Viral and Bacterial Communities of Colorectal Cancer

### 2017-03-17 doc/manuscript.Rmd
Huge update.

---
title: The Colorectal Cancer Virome
author: 
- name: Geoffrey D Hannigan
  affiliation: 1
- name: Melissa B Duhaime
  affiliation: 2
- name: Mack T Ruffin IV
  affiliation: 3
- name: Charlie C Koumpouras
  affiliation: 1
- name: Patrick D Schloss
  affiliation: 1,\*
address:
  - code: 1
    address: Department of Microbiology & Immunology, University of Michigan, Ann Arbor, Michigan, 48108
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48108
  - code: 3
    address: Department of Family and Community Medicine, Pennsylvania State University Hershey Medical Center, Hershey, Pennsylvania, 17033
  - code: \*
    address: To whom correspondence may be addressed.
output:
  md_document:
    variant: markdown
---

@@

***Text Length***: 34,182 / ~39,000 Characters

@@

Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world. Although the majority of colorectal cancer case causes remains unclear, it's progression has been linked to colonic bacterial community composition. Viruses are another important component of the colonic microbial community that have yet to be studied in colorectal cancer, despite their oncogenic potential. We evaluated the colorectal cancer virome (virus community) using a cohort of 90 human subjects with either healthy, adenomatous (precancerous), or cancerous colons. We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing methods to compare the colorectal cancer virome to the bacterial community. We found that alpha and beta diversity metrics were insufficient for detecting virome changes in colorectal cancer, but more sophisticated virome-based random forest models identified significant changes in the virus community. The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting the community was indirectly linked to colorectal cancer by modulating bacterial community structure and functionality. These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities. Together our findings add to the existing model for the role the microbiome plays in colorectal cancer development, thus providing us with a more complete understanding of colorectal cancer etiology.

**Word Count**: 226

# Significance Statement
Colorectal cancer is a leading cause of cancer-related death in the United States and worldwide. Its progression and severity have been linked to colonic bacterial community composition. Little is known about cancer-associated colon virus communities and their influence on bacteria. We began addressing this knowledge gap by identifying changes in colonic virus communities in colorectal cancer patients, and how they compare to bacterial community changes. The altered colorectal cancer virome was used in virome-based machine learning models to accurately classify patient cancer status. The cancer-virus classifier was driven primarily by bacteriophages (bacterial viruses). Wider phage host ranges were more strongly associated with colorectal cancer. The results suggest an indirect role for the virome impacting colorectal cancer by modulating their associated bacterial community.

**Word Count**: 125

@@

# Introduction
Due to their mutagenic abilities and their propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and because bacteria have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb], bacteriophages have the potential to indirectly impact cancer. The gut virome (the virus community of the gut) therefore has the potential to impact health and disease (e.g. cancer), and has been associated with diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. The precedence of the virome impacting human health and the strong association of bacterial communities with colorectal cancer suggest that colorectal cancer may be linked to altered virus communities.

Colorectal cancer is the second leading cause of cancer-related deaths in the United States [@Siegel:2014jo]. The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016, and over 500,000 Americans will have died from the disease [@Siegel:2014jo]. Although the majority of colorectal cancer case causes remains unclear, variation in colorectal bacterial communities have been linked to the disease [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb]. This work has led to the proposal of disease models in which bacteria colonize the colon, develop biofilms, promote infiltration, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu]. This association has also allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja]. While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and understanding underlying etiology, bacteria are only a subset of the colon microbiome. Viruses are another important component of the colon microbial community that have yet to be studied in the context of colorectal cancer. We evaluated colorectal cancer disruptions in virus and bacterial community composition in a human cohort whose stool was sampled at the three relevant stages of cancer developement: healthy, adenomatous, and cancerous.

Colorectal cancer is a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e. adenoma) in the large intestine [@Fearon:2011go]. If left untreated, the adenoma will develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death. Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr], and survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz]. Thus work such as this, which aims to power early detection and prevention of progression from early cancer stages, has a great potential to inform therapeutic development.

In this study we report that colorectal cancer is associated with variation in colonic virus communities. The majority of viruses identified within the virome were temperate bacteriophages. Just as the association between the bacterial community and colorectal cancer was driven by select influential bacteria including *Fusobacterium*, the association between the virome and colorectal cancer was also driven by a subset of influential phages. Our data suggest that the influential phages do not exclusively infect influential bacterial (e.g. *Fusobacterium* phages infecting the *Fusobacterium*), but rather act through the community as a whole. The implications of these findings are threefold. *First*, this supports a biological role for the virome in colorectal cancer development and suggests that more than bacteria are involved in the process. *Second*, we present a supplementary, or even alternative virus-based approach for classification modeling of colorectal cancer using stool samples. *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer.

@@

## The Colorectal Cancer Virome Cohort
The study cohort consisted of 90 human subjects, 30 of which had healthy colons, 30 of which had adenomas, and 30 which had carcinomas **(Figure \ref{sampleproc})**. Half of the stool was used to sequence the bacterial communities using both 16S rRNA gene and shotgun sequencing techniques. The other half of the stool samples were purified for virus like particles (VLPs) and consequent genomic DNA extraction, followed by shotgun metagenomic sequencing. The VLP purification allowed us to observe the *active virome* because we only sequenced those viruses that are encapsulated.

Virus DNA was purified before sequencing, which allowed us to analyze DNA from within virus capsids **(Figure \ref{sampleproc})**. Each extraction protocol was performed with a blank control to detect any contaminants from reagents. Only one of the nine controls contained detectable DNA, which was of a minimal concentration, thus providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**. As was expected, these controls were sparsely sequenced and were mostly removed while sub-sampling to even depths **(Figure \ref{qualcontrol} B)**. The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1kb **(Figure \ref{contigqc})**. Because contigs only represent genome fragments, we further clustered related contigs into operational genomic units (OGUs; conceptually similar to 16S rRNA gene operational taxonomic units) with the majority containing hundreds of related contigs **(Figure \ref{contigqc} - \ref{clustercontigqc})**.

## Virus Diversity and Colorectal Cancer
Microbiome and disease associations are often described as being of an altered diversity (i.e. "dysbiotic"). We therefore initially evaluated the diversity of virome OGUs and their association with colorectal cancer. We utilized the Bray-Curtis dissimilarity metric to evaluate differences in communities between disease states. To control for uneven sequencing depths, we subsampled to a minimum depth that maintained most samples while excluding the sparsely sequenced blank controls.

We calculated the differences in virus alpha diversity associated with colorectal cancer. We found no significant alterations in either Shannon entropy or richness **(Figure \ref{betaogu} C-D)**. There was also no statistically significant clustering of the disease groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogu}**). It is worth noting that there was a significant difference between the blank controls (those few that remained after sub-sampling) and the other study groups, further supporting the quality of our sample set (Anosim p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogunegative})**. Overall, standard diversity metrics were insufficient for capturing the differences in the virus communities between disease states.

## Virome Composition Changes in Colorectal Cancer
Previous work has shown that 16S rRNA gene relative abundance profiles are effective features for classifying stool samples as originating from healthy, adenomatous, or cancerous individuals [@Zackular:2014fba; @Baxter:2016dja]. The exceptional performance of bacteria in these classification models supports a role for bacteria in colorectal cancer. Here we built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared performance to models built using bacterial community signatures.

We built and tested random forest models to classify stool samples as belonging to either cancerous or healthy individuals. These models were based on virus metagenomic community or bacterial 16S rRNA gene relative abundance profiles. We confirmed that our bacterial 16S rRNA gene model replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba]. We then compared the bacterial 16S rRNA gene model to a model built using virome relative abundance. The viral model performed as well as the bacterial model (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = sigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = sigfig)`, respectively **(Figure \ref{predmodel} A - B)**.

To evaluate the synergistic capabilities of the bacteria and viruses within the model, we built a combinatory model that used both bacterial and viral community data. The combination model yielded modest but significantly improved performance beyond the separate virome (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Combined"),"pval"], digits = sigfig)`) and bacterial (corrected p-value = `r signif(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) models, and yielded an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**. This suggested that features from the virus and bacterial communities may have had synergistic capabilities for classifying stool as belonging to cancerous individuals.

We also compared these models to relative abundance profiles from bacterial metagenomic shotgun sequencing data. This model performed poorly, with a mean AUC of `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**. Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and lowly abundant OTUs **(Figure \ref{16scompare})**. Filtration of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome **(Figure \ref{16scompare} A)**. The majority of these OTUs had a relative abundance lower than 1% **(Figure \ref{16scompare} B)**.

@@

## Phage Influence Shifts During CRC Progression
Because our cohort included healthy, adenomatous, and cancerous colons, we were able to gain insight into virus community shifts during cancer progression. We evaluated community shifts between the two disease transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the sample classes around the transitions. We found that, while bacterial 16S rRNA gene models perform equally well for all disease class comparisons, the virome models performed worse when seperating healthy from adenomatous, and better when seperating adenomatous from cancerous **(Figure \ref{transitionmodels} A-B)**. Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members drove the transitions from healthy to adenomatous and adenomatous to cancerous, with one phage exception **(Figure \ref{transitionmodels} C-E)**. A *Myoviridae* OGU (Cluster 188) was the seventh most important phage driving the transition from a healthy to adenomatous state, and was the most important phage driving the progression from an adenomatous to carcingenic state. Therefore, like bacteria, there are distinct phages assocaited with the transitions from a healthy to an adenomatous colon, and from an adenomatous to a cancerous colon.

## Inclusive Colonic Virome Classification
After evaluating our ability to classify samples between two disease states, we performed a comprehensive three-class classification of all three disease states. We used a three-class random forest model for the bacterial 16S rRNA gene and viral sample sets. The bacterial signature model yielded an AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = sigfig)` and outperformed the viral community model which yielded an overall AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = sigfig)` (p-value = `r signif(threesig, digits = sigfig)`, **Figure \ref{threewaymodel} A-C**). Both models were best able to classify cancer samples from healthy or precancerous samples, but struggled to distinguish adenomatous from healthy or cancerous **(Figure \ref{threewaymodel} A-B)**. The cancerous signal was the most discriminatory of the three sample types.

The microbes important for the cancer vs healthy and healthy vs adenoma bacteria and virus models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**. The most important bacterium was the same *Fusobacterium* between the two and three class models, supporting its significance to the association between cancer and the bacterial communities **(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)**. The viruses most important to the three-class model were identified as bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**.

The classification model determined cancer state by incorporating the relative abundance profiles of the microbes within each community. The signatures ranged from notably high abundance of some OGUs, low abundance of some OGUs, and an absence of other OGUs **(Figure \ref{threewaymodel} F)**. Not all important OGUs were of increased abundance. The viral classification model depended on the unique signatures of these different abundance profiles to accurately classify each sample.

## Dominance of Bacteriophages in Colorectal Cancer Virome
Changes in the colorectal cancer virome could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts. To better understand the types of viruses that are important for colorectal cancer, we identified the virome OGUs as being similar to either eukaryotic viruses or bacteriophages. The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**. Overall we were able to identify `r signif(idcount, digits = sigfig)`% of the OGUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OGUs aligned to bacteriophage reference genomes.

We evaluated whether the phages in the community were primarily lytic (replicate by lysing their host) or temperate (lysogenic; able to integrate into their host's genome, as well as lyse the cell). We accomplished this by identifying three markers for temperate phages in the OGU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes. This approach was done as described in previous work [@Minot:2011ez; @Hannigan:2015fz]. We found that the majority of the colon phages were temperate, and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**. Thus the majority of the OGUs are temperate bacteriophages and not eukaryotic viruses, indicating the association between the virome and colorectal cancer is reliant on bacteriophage communities that can lie dormant in bacterial genomes. These findings are consistent with previous reports suggesting the gut virome is primarily temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].

## Community Context of Influential Phages
Because the link between colorectal cancer and the virome was driven by bacteriophages, we hypothesized that the influential phages were predators of the influential bacteria, and thus influenced their relative abundance through predation. If this hypothesis were true, we would expect the relative abundance between important bacteria and phages to be correlated. Instead we observed a strikingly low correlation between the bacterial and phage relative abundances **(Figure \ref{correlations} A,C)**. There was an overall absence of correlations between the most influential phage OGUs and bacterial OTUs **(Figure \ref{correlations} B)**. This evidence supported our null hypothesis that the influential phages are not primarily predators of influential bacteria.

>>>>

---
title: Viral and Bacterial Communities of Colorectal Cancer
author: 
- name: Geoffrey D Hannigan
  affiliation: 1
- name: Melissa B Duhaime
  affiliation: 2
- name: Mack T Ruffin IV
  affiliation: 3
- name: Charlie C Koumpouras
  affiliation: 1
- name: Patrick D Schloss
  affiliation: 1,\*
address:
  - code: 1
    address: Department of Microbiology & Immunology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48109
  - code: 3
    address: Department of Family and Community Medicine, Pennsylvania State University Hershey Medical Center, Hershey, Pennsylvania, 17033
  - code: \*
    address: To whom correspondence may be addressed.
output:
  md_document:
    variant: markdown
fontsize: 11pt
mainfont: 'Helvetica'
---

@@

@@

Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world. Although the majority of colorectal cancer case causes remain unclear, its progression has been linked to changes in colonic bacterial community composition. Viruses are another important component of the colonic microbial community, but have yet to be studied in colorectal cancer, despite their oncogenic potential. We evaluated the colorectal cancer virome (virus community) using a cohort of 90 human subjects with either healthy, adenomatous (precancerous), or cancerous colons. We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing methods to compare the colorectal cancer virome to the bacterial community. We found that alpha and beta diversity metrics were insufficient for detecting virome changes in colorectal cancer, but more sophisticated virome-based random forest models identified striking changes in the virus community. The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting the community was indirectly linked to colorectal cancer by modulating bacterial community structure and functionality. These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities. Together our findings add to the existing model for the microbiome's role in colorectal cancer development, thus providing a more complete understanding of colorectal cancer etiology.

**Word Count**: 226 / 250

# Significance Statement
Colorectal cancer is a leading cause of cancer-related death in the United States and worldwide. Its progression and severity have been linked to colonic bacterial community composition. Although viruses have been linked to other cancers and diseases, little is known about colorectal cancer virus communities. We began addressing this knowledge gap by identifying changes in colonic virus communities in colorectal cancer patients, and how they compared to bacterial community changes. The results suggested an indirect role for the virome impacting colorectal cancer by modulating their associated bacterial community. These findings both support a biological role for viruses in colorectal cancer, as well as provide a new understanding of basic colorectal cancer etiology.

**Word Count**: 113 / 120

@@

# Introduction
Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and because bacteria have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb], bacteriophages have the potential to indirectly impact cancer. The gut virome (the virus community of the gut) therefore has the potential to impact health and disease (e.g. cancer), and has already been associated with diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. The precedence of the virome impacting human health and the strong association of bacterial communities with colorectal cancer together suggest that colorectal cancer may be associated with altered virus communities.

Colorectal cancer is the second leading cause of cancer-related deaths in the United States [@Siegel:2014jo]. The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016, and over 500,000 Americans will have died from the disease [@Siegel:2014jo]. Although the majority of colorectal cancer case causes remain unclear, variation in colorectal bacterial communities have been linked to the disease [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb]. This work has led to a proposed disease model in which bacteria colonize the colon, develop biofilms, promote infiltration, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu]. This association has also allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja]. While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and understanding underlying etiology, bacteria are only a subset of the colon microbiome. Viruses are another important component of the colon microbial community that have yet to be studied in the context of colorectal cancer. We evaluated disruptions in virus and bacterial community composition in a human cohort whose stool was sampled at the three relevant stages of cancer development: healthy, adenomatous, and cancerous.

Colorectal cancer is a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e. adenoma) in the large intestine [@Fearon:2011go]. If left untreated, the adenoma will develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death. Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr], and survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz]. Thus work that aims to facilitate early detection and prevention of progression beyond early cancer stages has great potential to inform therapeutic development.

Here we begin addressing the knowledge gap in whether virus communities are altered in colorectal cancer and, if they are, how those changes might impact cancer progression and severity. We also aim to evaluate their diagnostic potential. We report that colorectal cancer is associated with variation in colonic virus community composition. The majority of viruses identified within the virome were temperate bacteriophages. Just as the association between the bacterial community and colorectal cancer was driven by select influential bacteria including *Fusobacterium*, the association between the virome and colorectal cancer was driven by a subset of influential phages. Our data suggest that the influential phages do not exclusively infect influential bacterial, but rather act through the community as a whole. The implications of these findings are threefold. *First*, this supports a biological role for the virome in colorectal cancer development and suggests that more than bacteria are involved in the process. *Second*, we present a supplementary, or even alternative virus-based approach for classification modeling of colorectal cancer using stool samples. *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer.

@@

## Cohort Design, Sample Collection, and Processing
The study cohort consisted of 90 human subjects, 30 of which had healthy colons, 30 of which had adenomas, and 30 which had carcinomas **(Figure \ref{sampleproc})**. Half of the stool was used to sequence the bacterial communities using both 16S rRNA gene and shotgun sequencing techniques. The other half of the stool samples were purified for virus like particles (VLPs) before genomic DNA extraction and shotgun metagenomic sequencing. The VLP purification allowed us to observe the *active virome* because we only sequenced those viruses that were encapsulated.

Virus DNA was purified prior to sequencing, allowing us to analyze DNA exclusively from within virus capsids **(Figure \ref{sampleproc})**. Each extraction protocol was performed with a blank control to detect contaminants from reagents or other unintentional sources. Only one of the nine controls contained detectable DNA, which was of a minimal concentration, thus providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**. As was expected, these controls were sparsely sequenced and were mostly removed while sub-sampling to even depths **(Figure \ref{qualcontrol} B)**. The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1kb **(Figure \ref{contigqc})**. Because contigs only represent genome fragments, we further clustered related contigs into operational genomic units (OGUs) **(Figure \ref{contigqc} - \ref{clustercontigqc})**.

## Unaltered Virome Diversity in Colorectal Cancer
Microbiome and disease associations are often described as being of an altered diversity (i.e. "dysbiotic"). We therefore initially evaluated the changes in virome OGU diversity in colorectal cancer. We evaluated differences in communities between disease states using the Bray-Curtis, Shannon entropy, and richness diversity metrics. To control for uneven sequencing depths, we subsampled to a minimum depth that maintained most samples while excluding the sparsely sequenced blank controls.

We observed no significant alterations in either Shannon entropy or richness in the diseased state **(Figure \ref{betaogu} C-D)**. There was also no statistically significant clustering of the disease groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogu}**). It is worth noting that there was a significant difference between the few blank controls that remained after subsampling, and the other study groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogunegative})**. This further supported the quality of our sample set. Overall, standard diversity metrics were insufficient for capturing virus community differences between disease states.

## Altered Virome Composition in Colorectal Cancer
16S rRNA gene relative abundance profiles are effective feature sets for classifying stool samples as originating from healthy, adenomatous, or cancerous individuals [@Zackular:2014fba; @Baxter:2016dja]. The exceptional performance of bacteria in these classification models supports a role for bacteria in colorectal cancer. We built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared performance to models built using bacterial community signatures.

To identify the altered virus communities associated with colorectal cancer, we built and tested random forest models to classify stool samples as belonging to either cancerous or healthy individuals. These models were based on virus metagenomic community relative abundance profiles and were compared to 16S rRNA gene models. We confirmed that our bacterial 16S rRNA gene model replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba]. We then compared the bacterial 16S rRNA gene model to a model built using virome relative abundance. Statistically, the viral model performed better than the bacterial model (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`), although the results were practically similar, with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = sigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = sigfig)`, respectively **(Figure \ref{predmodel} A - B)**.

To evaluate the synergistic capabilities of the bacteria and viruses, we built a combinatory model that used both bacterial and viral community data. The combination model yielded statistically significant but minor improved performance beyond the separate virome (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Combined"),"pval"], digits = sigfig)`) and bacterial (corrected p-value = `r signif(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**. This suggested that features from the virus and bacterial communities may have had synergistic capabilities for classifying stool as belonging to cancerous individuals.

To determine the advantage of viral metagenomic methods over bacterial metagenomic methods, we compared the viral model to a model built using relative abundance profiles from bacterial metagenomic shotgun sequencing data. This model performed more poorly than the viral model, with a mean AUC of `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**. This performance was also much poorer than that observed in the 16S rRNA gene model. Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and lowly abundant OTUs **(Figure \ref{16scompare})**. Filtration of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome **(Figure \ref{16scompare} A)**. The majority of these OTUs had a relative abundance lower than 1% **(Figure \ref{16scompare} B)**.

@@

## Shifted Phage Influence Across Cancer Stages
Because our cohort included healthy, adenomatous, and cancerous colons, we were able to evaluate shifts in influential phage identities during cancer progression. We evaluated community shifts between the two disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the sample classes around the transitions. While bacterial 16S rRNA gene models performed equally well for all disease class comparisons, the virome model performances differed **(Figure \ref{transitionmodels} A-B)**. Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members were important in the transitions from healthy to adenomatous and adenomatous to cancerous stages **(Figure \ref{transitionmodels} C-E)**.

After evaluating our ability to classify samples between two disease states, we performed a three-class random forest model including all disease states. The 16S rRNA gene model model yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = sigfig)` and outperformed the viral community model which yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = sigfig)` (p-value = `r signif(threesig, digits = sigfig)`, **Figure \ref{threewaymodel} A-C**).

The microbes important for the cancer vs healthy and healthy vs adenoma models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**. The most important bacterium was the same *Fusobacterium* between the two and three class models, supporting its significance to the association between cancer and the bacterial communities **(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)**. The viruses most important to the three-class model were also identified as bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**.

The classification model determined cancer state by incorporating the relative abundance profiles of the microbes within each community. Influential phage relative abundances were both higher and lower in cancer **(Figure \ref{threewaymodel} F)**. Not all important OGUs were of increased abundance in the diseased state. The viral classification model depended on the unique signatures of these different abundance profiles to accurately classify each sample.

## Bacteriophage Dominance in Colorectal Cancer Virome
Changes in the colorectal cancer virome could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts. To better understand the types of viruses that are important for colorectal cancer, we identified the virome OGUs as being similar to either eukaryotic viruses or bacteriophages. The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**. Overall we were able to identify `r signif(idcount, digits = sigfig)`% of the OGUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OGUs aligned to bacteriophage reference genomes.

We evaluated whether the phages in the community were primarily lytic (replicate by lysing their hosts) or temperate (lysogenic; able to integrate into their host's genome, as well as lyse the cell). We accomplished this by identifying three markers for temperate phages in the OGU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes, as previously described [@Minot:2011ez; @Hannigan:2015fz]. We found that the majority of the colon phages were temperate, and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**. Thus the majority of the OGUs are temperate bacteriophages and not eukaryotic viruses, indicating the association between the virome and colorectal cancer is reliant on bacteriophage communities that can lie dormant in bacterial genomes. These findings are consistent with previous reports suggesting the gut virome is primarily temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].

## Community Context of Influential Phages
Because the link between colorectal cancer and the virome was driven by bacteriophages, we hypothesized that the influential phages were primarily predators of the influential bacteria, and thus influenced their relative abundance through predation. If this hypothesis were true, we would expect a correlation between the relative abundances of influential bacteria and phages. Instead we observed a strikingly low correlation between the bacterial and phage relative abundances **(Figure \ref{correlations} A,C)**. There was an overall absence of correlation between the most influential phage OGUs and bacterial OTUs **(Figure \ref{correlations} B)**. This evidence supported our null hypothesis that the influential phages were not primarily predators of influential bacteria.

Given these findings, we hypothesized that the influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria. This hypothesis would be supported by showing that the influential bacteriophages were community hubs within the bacteria and phage interactive network. We investigated the infectious capabilities and potential host ranges of all phage OGUs using a random forest model to predict which phages infected which bacteria in the overall community (**Cite Network Preprint**). The predicted interactions were then used to identify community hub phages. This revealed a wide tropism range for the bacteriophages within the community **(Figure \ref{network} A)**. We calculated the alpha centrality (measure of importance in the ecosystem network) of each phage OGU's connection to the rest of the network, and compared the centrality to the importance of each OGU in the colorectal cancer classification model. The phages with high centrality values were defined as community hubs. We found that phage OGU centrality is significantly positively correlated with importance to the disease model (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer were also more likely to be community hubs **(Figure \ref{network} B)**. Together these findings supported our hypothesis that influential phages were hubs within their microbial communities.

# Discussion
Because of their propensity for mutagenesis and capacity for modulating their host functionality, many viruses are oncogenic [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Some bacteria also have oncogenic properties, meaning bacteriophages may play an indirect role in promoting carcinogenesis by influencing bacterial communities [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb]. Despite their carcinogenic potential and the strong association between bacteria and colorectal cancer, the link between virus colorectal communities and colorectal cancer has yet to be evaluated. Here we show that, like colonic bacterial communities, the colon virome was altered in colorectal cancer. Our findings support a working hypothesis for oncogenesis by phage-modulated bacterial community composition.
